---
title: "Reproducible Research: Peer Assessment 1"
author: "Imran Saeed"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Questions to be answered:

* What is mean total number of steps taken per day?
* What is the average daily activity pattern?
* Imputing missing values
* Are there differences in activity patterns between weekdays and weekends?

##  Installation and setup..
###Installing and lodaing all the packages needed for this assignment. Will install only if not installed. 
```{r package setup and installation}
listOfPackages <- c("ggplot2","plyr")
newPackages <- listOfPackages[!(listOfPackages %in% installed.packages()[,"Package"])]

if(length(newPackages)) {
	message(sprintf("Going to install package(s)[ %s ]\n", newPackages))
	install.packages(newPackages) 
} else {
	message("All packages are already installed. Will skip install pahse")
}


## Load all of the package needed for this project.
for(package in listOfPackages){
  library(package, character.only = TRUE)
}
```


## Loading and preprocessing the data

#### Generic variables 

```{r Generic variables}

zipDataFileName <- "repdata%2Fdata%2Factivity.zip"
dataFileDir <- "repdata%2Fdata%2Factivity"
zipDataFilePath <- paste0(getwd(),"/", zipDataFileName)
dataFileName <- "activity.csv"
dataFilePath <- paste0(getwd(),"/", dataFileName)
```

#### 1.1 Load the dataset
```{r load data}
if (!file.exists(zipDataFilePath)){
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, zipDataFilePath, method="curl")
}
#### STEP 1.1 - Unzip previously obtained data
if (file.exists(zipDataFilePath) & !file.exists(dataFileDir)) { 
  unzip(zipDataFilePath) 
}

#### STEP 1.2 - Read the CSV data
activityDataSet <- read.csv(dataFileName, sep = ",")

```
The variable names and the structure of the file are given by
```{r activity details}

names(activityDataSet)
str(activityDataSet)


activityDataSet$dateTime<- as.POSIXct(activityDataSet$date,  format="%Y-%m-%d")

## clean data set where steps are not NAs
cleanDataSet <- activityDataSet[!is.na(activityDataSet$steps),]
head(cleanDataSet)

```

## What is mean total number of steps taken per day?
```{r activity steps per day}
stepsPerDay <- aggregate(steps ~ date, activityDataSet, FUN = sum, na.rm=TRUE)

ggplot(stepsPerDay, aes(steps)) + geom_histogram(bins = 30) 

```

Summery of data set.
```{r activity summary}
summary(stepsPerDay$steps)

```


Mean of steps.
```{r steps mean}
mean(activityDataSet$steps)
```

Median of steps.
```{r steps median}
median(activityDataSet$steps)

```

## What is the average daily activity pattern?

```{r daily activity pattern}
meanActivity <- aggregate(steps ~ interval , activityDataSet , mean)
ggplot(meanActivity, aes(x=interval, y = steps)) + geom_line()

```

 Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?

```{r Maximum steps in  5-minute interval}

arrange(meanActivity, -steps)[1,]

```

## Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the data set (i.e. the total number of rows with NAs)

```{r Total number of missing values}

sum( is.na(activityDataSet$steps ))

```

Devise a strategy for filling in all of the missing values in the data set. 

```{r Strategy for filling in all of the missing values}

activityDataSetCopy <- activityDataSet
meanInterval <- aggregate(steps~interval, activityDataSet, mean)

```

Create a new data set that is equal to the original data set but with the missing data filled in.

```{r Fill missing values with mean}

for(i in 1:nrow(activityDataSetCopy)){
  if(is.na(activityDataSetCopy[i,1])){
     rowNo <- which(meanInterval$interval==activityDataSetCopy[i,]$interval)
    activityDataSetCopy[i,1] <- meanInterval[ rowNo, ]$steps
  }
}

```   

Histogram of the total number of steps taken each day after imputing missing values

```{r Histogram of the total number of steps taken each day after imputing missing values}

sumSteps2 <- aggregate(steps~date, data= activityDataSetCopy, sum)
ggplot(sumSteps2, aes(steps)) + geom_histogram(bins=30) 

```


Mean after imputing missing values

```{r Mean after imputing missing values}

mean(sumSteps2$steps)

```

Median after imputing missing values

```{r Median after imputing missing values}

median(sumSteps2$steps)

```

## Are there differences in activity patterns between weekdays and weekends?


```{r preparing data set to plot after imputing NAs, echo=TRUE}
activityDataSet2 <- activityDataSet
activityDataSet2$day <- weekdays(   as.Date(activityDataSet2$date) )
activityDataSet2$day2[activityDataSet2$day=="Saturday" | activityDataSet2$day=="Sunday"] <- "Weekend"
activityDataSet2$day2[is.na(activityDataSet2$day2) ] <- "Weekday"

activityDataSet2 <- aggregate(steps~interval + day2, activityDataSet2, mean)
```


```{r Total number of steps taken each day after imputing values}

ggplot(activityDataSet2,aes(x=interval, y=steps)) + facet_grid(day2~.) + geom_line()

```
